---
title: "lecture01-intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lecture01-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to the course "Appelied Spatial Data Analysis and Research"
course details: https://opas.peppi.uef.fi/en/course/YH00EM30/135183?period=2025-2026

## What is R?

### Why should we learn R?
R follows a type inference coding structure and provides a wide variety of statistical and graphical techniques, including;

- Linear and non-linear modelling
- Univariate & Multivariate Statistics
- Classical statistical tests
- Time-series analysis/ Econometrics
- Simulation and Modelling
- Datamining-classification, clustering etc.

For computationally intensive tasks, C, C++, and Fortran code can be linked and called at run time.

R is easily extensible through functions and extensions, and the R community is noted for its active contributions in terms of packages.

```{r}
# Number of R Packages
length(available.packages(repos = "http://cran.us.r-project.org")[, 1])
```

### Installing R and RStudio on Windows

The latest version of R can be download from the R homepage.

R download page: http://www.cran.r-project.org/bin/windows/base/ The page also provides some instructions and FAQ’s on R installation.

RStudio IDE ( IDE: Integrated Development Environment) is a powerful and productive user interface for R.

It’s free and open source, and works great on Windows, Mac, and Linux

### RStudio GUI/IDE

RStudio GUI is composed of 4 panes which can be rearranged according to the requirements.

There are a lot of short introductions to RStudio available online so we will not go into more details.

Download Rstudio from here https://rstudio.com/products/rstudio/download/#download

### Installing Packages

The easiest way to install packages is to do it via R console. The command install.packages(“package name”) installs R packages directly from internet. Other options to install various dependencies to a package can be easily specified when calling this function. A call to this function asks the user to chose a CRAN mirror at the first instance.

Run the following to install Quantreg package on R. Also use the help function to get the details.

```{r}
# Install a package using RStudio Console
install.packages("sf", dependencies = c("Depends", "Suggests"))
```

```{r}
install.packages(c("reshape2", "foreign", "ggplot2", "stargazer"), dependencies = TRUE)
# to be updated
```

### Getting Help

As R is constantly evolving and new functions/packages are introduced every day it is good to know sources of help. The most basic help one can get is via the help() function. This function shows the help file for a function which has been created by package managers.

```{r}
help("function name")
```
    
All the R packages (with few exceptions) have a user’s manual listing the functions in a package. This can be downloaded in PDF format from the R package download page2.

R also provides some search tools given at http://cran.r-project.org/search.html The R Site search is helpful in searching for topics related to problem in hand.

Other than these various good R related blogs are on the internet which can be really helpful. A combined upto date view of 452 contributed blogs can be found at R-bloggers3.

Over all there quite a big community of R Users and help can be found for most of the topics.


## Reading data

###Reading data

---
title: "Open Data: Statistics Finland WFS service"
output: html_document
---

## Introduction

This example demonstrates how to download, explore, and process open spatial data from the Statistics Finland WFS service.  
We will:

1. Inspect the WFS service  
2. Download a 5 km grid dataset  
3. Visualize the grid using **ggplot2**  
4. Clip the grid to the municipality of Kotka  
5. Save the clipped dataset as a shapefile  

This workflow helps you understand how openly available geospatial data can be accessed and integrated into spatial analysis.

---

## 1. Load Required Libraries

In this step, we load the R packages used throughout the analysis. These include tools for spatial data handling (sf), data manipulation (dplyr, purrr), downloading data from web services (httr, ows4R), and visualization (ggplot2). The geofi package is also loaded to access official Finnish municipal boundary data.

```{r}
library(dplyr)
library(purrr)
library(sf)
library(httr)
library(data.table)
library(ows4R)
library(ggplot2)
library(geofi)   # for Finnish municipal boundaries
```


## 2. Inspect the WFS Service
Statistics Finland provides geospatial datasets through a WFS service.

Before downloading any data, we connect to the Statistics Finland Web Feature Service (WFS). Using the WFSClient from the ows4R package, we query the service to list all available datasets. This ensures we know which layers (e.g., grids, municipality borders, zip codes) can be accessed and what their names are.
```{r}
vayla <- "https://geo.stat.fi/geoserver/tilastointialueet/wfs"

vayla_client <- WFSClient$new(
  url = vayla,
  serviceVersion = "2.0.0"
)

vayla_client$getFeatureTypes(pretty = TRUE)
```

### 3. Download the 5 km Grid Dataset

Once we know which dataset we want, we construct a WFS request URL that retrieves the 5 km statistical grid (called hila5km) in GeoJSON format. The st_read() function from sf is then used to download and immediately load the spatial data into R as an sf object.

So, let´s construct a WFS request and load the data directly into an sf object.
```{r}
url_grid <- list(
  hostname = "geo.stat.fi/geoserver/tilastointialueet/wfs",
  scheme   = "https",
  query = list(
    service = "WFS",
    version = "2.0.0",
    request = "GetFeature",
    typename = "tilastointialueet:hila5km",
    outputFormat = "json"
  )
) %>% setattr("class", "url")

grid_url <- build_url(url_grid)

grid_5km <- st_read(grid_url)
```

### 4. Visualize the Grid

In this step, we create a simple map of the downloaded 5 km grid. The ggplot2 and geom_sf() functions allow us to quickly inspect the geometry and structure of the grid, ensuring the data loaded correctly and looks as expected.

```{r}
ggplot(grid_5km) +
  geom_sf(aes(fill = id), color = NA) +
  labs(title = "Statistics Finland 5 km Grid")
```

### 5. Clip the Grid to Kotka

To focus the analysis on a specific municipality, we first download Finland’s municipal boundaries using geofi::get_municipalities(). We then compute the centroid of each grid cell and perform a spatial join to determine which grid cells fall inside the municipality of Kotka. This isolates only the portion of the grid that overlaps Kotka’s area.

5.1 Retrieve municipality boundaries with geofi-package

The geofi package provides streamlined access to Finnish administrative and statistical geospatial data through the Statistics Finland WFS API, making it easy for researchers and analysts to obtain harmonized spatial datasets for diverse applications ranging from urban planning to environmental research.  In addition to spatial layers such as municipalities, postal code areas, and population grids, geofi includes annually updated municipality key tables that support regional aggregation and multilingual data handling. 

Study the geofi package website, https://ropengov.github.io/geofi/, carefully, as there will be an exercise based on it this week.

Let´s now download data from the municipalities by using geofi-package

```{r}
municipalities <- geofi::get_municipalities(year = 2022) %>%
  select(kunta, kunta_name)
```

5.2 Generate centroids for each grid cell
```{r}
grid_centroids <- st_centroid(grid_5km)
```

5.3 Select grid cells located in Kotka

```{r}
kotka <- subset(municipalities, kunta_name == "Kotka")

grid_in_kotka <- st_join(grid_centroids, kotka)

kotka_grid <- subset(grid_in_kotka, kunta_name == "Kotka")
```

Visualize the results
```{r}
ggplot(kotka_grid) +
  geom_sf(color = "red") +
  labs(title = "Statistics Finland 5 km Grid in Kotka")
```

### 6. Export the Result as a Shapefile

Finally, we save the clipped grid as a shapefile using st_write(). Shapefiles can be opened in GIS software such as QGIS or ArcGIS, making this step useful for further spatial analysis, map production, or sharing results with others.

```{r, eval=FALSE}
st_write(kotka_grid,
         "define your path here/r1km_kotka.shp")
```





```{r setup}
library(spatialcourseOL)
usethis::use_pkgdown()

```
